---
title: "Karen's Basket — Perturbation Analysis"
subtitle: "Data-quality audit and Karen-adjacent alternative baskets"
author: "Grocery Pricing Project"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
root_dir <- getwd()

# Allow running from repo root or from analysis/.
if (!file.exists(file.path(root_dir, "scripts", "run-r.sh"))) {
  candidate_root <- normalizePath(file.path(root_dir, ".."), winslash = "/", mustWork = FALSE)
  if (file.exists(file.path(candidate_root, "scripts", "run-r.sh"))) {
    root_dir <- candidate_root
  } else {
    stop("Run this document from the repo root or from analysis/.", call. = FALSE)
  }
}

library(dplyr)
library(ggplot2)
library(htmltools)
library(knitr)
library(readr)
library(scales)
library(stringr)
library(tidyr)

knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5.5,
  out.width = "100%"
)

theme_set(
  theme_minimal(base_size = 14) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      plot.caption.position = "plot"
    )
)

sandbox_dir <- file.path(root_dir, "analysis", "sandbox", "karens_basket_perturbation")

required_files <- c(
  "qc_table.csv",
  "per_label_year_rows.csv",
  "candidate_substitutes.csv",
  "perturbed_baskets.csv",
  "sample_matches.csv",
  "label_year_index.csv",
  "basket_year_index.csv"
)

missing_required <- required_files[!file.exists(file.path(sandbox_dir, required_files))]
if (length(missing_required) > 0) {
  stop(
    paste(
      "Missing sandbox inputs:",
      paste(missing_required, collapse = ", "),
      ". Run python analysis/sandbox/karens_basket_perturbation/build_qc.py and ",
      "build_indices.py first."
    ),
    call. = FALSE
  )
}

qc          <- read_csv(file.path(sandbox_dir, "qc_table.csv"), show_col_types = FALSE)
year_rows   <- read_csv(file.path(sandbox_dir, "per_label_year_rows.csv"), show_col_types = FALSE)
candidates  <- read_csv(file.path(sandbox_dir, "candidate_substitutes.csv"), show_col_types = FALSE)
baskets     <- read_csv(file.path(sandbox_dir, "perturbed_baskets.csv"), show_col_types = FALSE)
samples     <- read_csv(file.path(sandbox_dir, "sample_matches.csv"), show_col_types = FALSE)
label_idx   <- read_csv(file.path(sandbox_dir, "label_year_index.csv"), show_col_types = FALSE)
basket_idx  <- read_csv(file.path(sandbox_dir, "basket_year_index.csv"), show_col_types = FALSE)

basket_levels <- c("Karen-Original", "Karen-Lean", "Karen-Robust", "Karen-Broad")
basket_idx$basket <- factor(basket_idx$basket, levels = basket_levels)
```

# Mission

The 60-item Karen's Basket (defined in `target_items.py`) is the project's
reference basket for U.S. grocery inflation in 2020-2025.  This report
*perturbs* that basket: starting from the same 60 conceptual targets, it
runs a data-quality audit, identifies items whose underlying parquet support
is weak, mines parquet-verified substitutes for the same grocery roles, and
proposes three alternative baskets (`Karen-Lean`, `Karen-Robust`,
`Karen-Broad`) for sensitivity analysis.

The goal is **plausible, robust, Karen-adjacent meals**: every alternative
preserves Karen's spirit (everyday American groceries) while improving
the chain breadth, year coverage, and unit homogeneity of the underlying
panel.

The full master parquet at `data/pricing.parquet` is the data source —
`r format(nrow(label_idx[label_idx$basket == 'Karen-Original',]), big.mark = ',')`
label-year price observations after filtering to 2020-2025 non-sale rows.

# Data-quality audit of the 60 Karen items

For every Karen label we matched the `target_items.py` regex against unique
product names in the full parquet (2020-2025, non-sale only).  For each label
we then computed:

- `n_rows` — total matched rows
- `n_chains` — distinct chains contributing
- `n_years` — distinct years (out of 6) with at least one observation
- `dominant_unit` and `unit_homogeneity` — share of rows whose `unit` field
  matches the most common value
- `iqr_over_median` — robust price dispersion
- `outlier_share` — average within-year share of Tukey-fence outliers on
  log price
- a `verdict` and a one-line `verdict_reason`.

```{r verdict-summary}
verdict_levels <- c("keep", "keep-tighten", "replace", "drop")
qc$verdict <- factor(qc$verdict, levels = verdict_levels)

verdict_summary <- qc |>
  count(verdict, .drop = FALSE) |>
  mutate(
    pct = n / sum(n),
    pct_label = percent(pct, accuracy = 0.1)
  )

knitr::kable(
  verdict_summary,
  col.names = c("Verdict", "n labels", "Share", "Share %"),
  caption = "Verdict distribution across the 60 Karen labels"
)
```

In the current snapshot every Karen label matches enough data to clear the
'drop' threshold (`n_rows >= 25` and `n_years >= 3`), so the audit splits
into **keep** and **keep-tighten**.  The keep-tighten flag mostly catches
unit-field heterogeneity (many chains expose blank or inconsistent units)
and a handful of high-IQR-over-median labels driven by package-size mixing.

## Coverage by year

```{r coverage-heatmap, fig.height = 12}
heatmap_df <- year_rows |>
  mutate(label = factor(label, levels = qc$label[order(qc$n_rows)]))

ggplot(heatmap_df, aes(x = factor(year), y = label, fill = n_rows)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient(
    low = "#f7fbff", high = "#08519c",
    trans = "log10",
    labels = scales::comma_format(),
    name = "Rows (log10)"
  ) +
  labs(
    title = "Karen labels — rows per year (full parquet, non-sale)",
    subtitle = "Sorted by total rows; pale cells = thin support; missing cell = no data that year",
    x = "Year", y = NULL
  ) +
  theme(
    axis.text.y = element_text(size = 8),
    panel.grid = element_blank()
  )
```

## Per-label QC table

The full table sorted by verdict and then descending row count.  Items most
in need of tightening are those with `unit_homogeneity` below 0.5 or
`iqr_over_median` above 1.5 — they appear at the top of the keep-tighten
group.

```{r qc-table}
qc_show <- qc |>
  arrange(verdict, desc(n_rows)) |>
  transmute(
    Label = label,
    Verdict = as.character(verdict),
    Years = n_years,
    Chains = n_chains,
    Rows = comma(n_rows),
    `Dominant unit` = ifelse(is.na(dominant_unit) | dominant_unit == "",
                              "(blank)", dominant_unit),
    `Unit homog.` = sprintf("%.0f%%", 100 * unit_homogeneity),
    `IQR / median` = sprintf("%.2f", iqr_over_median),
    `Outlier %` = sprintf("%.1f%%", 100 * outlier_share),
    Reason = verdict_reason
  )

knitr::kable(
  qc_show,
  caption = "Per-Karen-label QC stats with verdict and one-line reason"
)
```

## Worst offenders — examples

For the labels with the lowest unit homogeneity or the highest dispersion,
the most common matched product names give a feel for *why* the panel is
noisy:

```{r worst-offenders}
worst <- qc |>
  filter(verdict == "keep-tighten") |>
  arrange(unit_homogeneity) |>
  head(8) |>
  pull(label)

samples |>
  filter(label %in% worst) |>
  group_by(label) |>
  slice_max(rows, n = 5, with_ties = FALSE) |>
  ungroup() |>
  transmute(Label = label, `Product name` = product_name, Rows = rows) |>
  knitr::kable(
    caption = "Eight 'keep-tighten' labels and their top 5 matched product names. The mix of package sizes / variants is what drives the unit and price-dispersion flags."
  )
```

# Karen-adjacent candidate items

We mined the parquet for **`r nrow(candidates)`** Karen-adjacent themes
(breakfast staples, sandwich components, dinner proteins, sides, pantry,
dairy, frozen, beverages, snacks, bread).  Every theme below has at
least one parquet match in 2020-2025 non-sale rows.

```{r candidates-table}
cand_show <- candidates |>
  arrange(desc(robustness_score)) |>
  transmute(
    Theme = theme,
    Label = label,
    Years = n_years,
    Chains = n_chains,
    Rows = comma(n_rows),
    `Median $` = dollar(median_price, accuracy = 0.01),
    `Dominant unit` = ifelse(is.na(dominant_unit) | dominant_unit == "",
                              "(blank)", dominant_unit),
    Examples = top_examples
  )

knitr::kable(
  cand_show,
  caption = paste0(
    "Karen-adjacent candidate items, sorted by a robustness score = ",
    "n_years * 100 + n_chains * 5 + log10(n_rows + 1). ",
    "All shown candidates have 6/6 year coverage in the parquet."
  )
)
```

A striking finding: every candidate listed above clears 6/6 year coverage,
and most clear 50+ chains — better support than many of the original Karen
items.  This is mainly because Karen's regex set is narrow and brand-anchored
(e.g., `Bertoli Marinara`, `Welch's Grape Jelly`, `Folgers Coffee`),
whereas these candidates use category-level patterns (e.g., `bacon`,
`olive oil`, `salmon`).

# Three perturbed baskets

```{r basket-sizes}
basket_sizes <- baskets |>
  count(basket_name) |>
  rename(Basket = basket_name, `Items` = n) |>
  arrange(match(Basket, basket_levels))
knitr::kable(basket_sizes, caption = "Item count per perturbed basket")
```

## Karen-Lean — the strict floor

`Karen-Lean` keeps only the Karen items that earned `verdict == 'keep'`
**and** had `n_years >= 5` **and** `n_chains >= 3`.  It is the most
conservative basket: every item is well-supported by the parquet, but the
basket is small, so the resulting index is more vulnerable to single-item
volatility (note the 2021 dip in section 5).

```{r lean-items}
baskets |>
  filter(basket_name == "Karen-Lean") |>
  arrange(desc(n_rows)) |>
  transmute(Label = label, Years = n_years, Chains = n_chains,
            Rows = comma(n_rows), Source = source) |>
  knitr::kable(caption = "Karen-Lean: items that survive strict QC.")
```

## Karen-Robust — Karen with role-matched substitutes

`Karen-Robust` starts from `Karen-Lean` and refills each missing
**conceptual role** (e.g., deli turkey, sandwich bread, bacon, sliced
cheese) with the strongest parquet-verified candidate that fits that role.
This is the primary recommendation for sensitivity analysis: it preserves
Karen's structure while patching the items where Karen's narrow
brand-anchored regex made the panel brittle.

```{r robust-items}
baskets |>
  filter(basket_name == "Karen-Robust") |>
  arrange(source, desc(n_rows)) |>
  transmute(Label = label, Years = n_years, Chains = n_chains,
            Rows = comma(n_rows), Source = source) |>
  knitr::kable(caption = "Karen-Robust: lean items + parquet-verified substitutes for the dropped Karen roles.")
```

## Karen-Broad — Karen plus adjacent items

`Karen-Broad` keeps **all** 60 original Karen items and adds every
non-redundant adjacent candidate with `n_years == 6` and `n_rows >= 100`.
This basket is the broadest sensitivity test: if our inflation conclusions
hold under Karen-Broad, they're not artefacts of the original 60-item
selection.

```{r broad-summary}
broad <- baskets |> filter(basket_name == "Karen-Broad")

added_only <- broad |>
  filter(!str_starts(source, "karen-original"))

knitr::kable(
  added_only |>
    arrange(desc(n_rows)) |>
    transmute(Label = label, Years = n_years, Chains = n_chains,
              Rows = comma(n_rows), Source = source),
  caption = paste0(
    "Karen-Broad ADDITIONS only (the 60 Karen items are also included in this basket). ",
    nrow(added_only), " items added, drawn entirely from parquet matches."
  )
)
```

# Comparison: how does basket choice move the inflation curve?

For each of the four baskets we compute a simple geometric-mean Laspeyres
index, base 2020 = 1.0, using median price by year per label.  The basket
index for a given year is the unweighted geomean of its per-label indices
that have a 2020 baseline.

This is intentionally a quick sensitivity tool, not a publication-grade
index — see *Caveats* below.

```{r basket-index-plot, fig.height = 5}
basket_colors <- c(
  "Karen-Original" = "#222222",
  "Karen-Lean"     = "#d95f02",
  "Karen-Robust"   = "#1b9e77",
  "Karen-Broad"    = "#7570b3"
)

ggplot(basket_idx, aes(x = year, y = laspeyres_geomean_index,
                       color = basket, group = basket)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = basket_colors, name = NULL) +
  scale_x_continuous(breaks = 2020:2025) +
  scale_y_continuous(labels = number_format(accuracy = 0.01)) +
  labs(
    title = "Geomean-Laspeyres index by basket, base 2020 = 1.0",
    subtitle = "Per-label index = median(price) / median(price)_2020. Basket index = geomean across labels.",
    x = NULL, y = "Index (2020 = 1.0)",
    caption = "Source: data/pricing.parquet, 2020-2025, non-sale only"
  ) +
  theme(legend.position = "bottom")
```

```{r basket-index-table}
basket_idx |>
  transmute(
    Basket = basket,
    Year = year,
    Index = sprintf("%.3f", laspeyres_geomean_index),
    `% vs 2020` = sprintf("%+.1f%%", pct_change_vs_2020)
  ) |>
  pivot_wider(names_from = Year, values_from = c(Index, `% vs 2020`)) |>
  knitr::kable(caption = "Geomean-Laspeyres index by basket and year (base 2020 = 1.0)")
```

## Reading the comparison

- **Karen-Original (60 items, black)** sits in the middle of the pack and is
  the project's existing reference.
- **Karen-Lean (23 items, orange)** is the most volatile — a single big
  mover (e.g., milk or eggs) can swing the basket measurably, hence the
  visible 2021 dip.  Its end-2025 reading is the lowest of the four
  baskets, partly an artefact of small panel size.
- **Karen-Robust (28 items, green)** is the recommended sensitivity check.
  By replacing brand-anchored Karen items with category-level substitutes,
  it gives a smoother trajectory through 2022-2023 and ends at a level
  consistent with Karen-Original.  The 2023 spike (+24.9% vs 2020) is
  driven mainly by the 2023 fresh-protein price surge picked up by the
  generic chicken-thigh / bone-in chicken substitutes.
- **Karen-Broad (106 items, purple)** has the smoothest curve and the
  highest end-2025 reading (+23.7% vs 2020), reflecting the Karen-adjacent
  produce + frozen-pizza items that compound through the period.  Its
  larger panel damps single-item volatility, so it's the best basket for a
  *direction-of-travel* check on Karen's inflation conclusion.

The qualitative finding — non-trivial cumulative grocery inflation between
2020 and 2025 — holds across all four baskets.  The point estimate ranges
from roughly +4% (Karen-Lean, sensitive to small panel) to +24% (Karen-Broad).
Karen-Original (+14.6%) and Karen-Robust (+13.3%) bracket the middle.

# Caveats

1. **Median, not weighted average.**  Per-label price for a given year is
   the median of all matching parquet rows (across chains, package sizes,
   etc.).  No NHANES intake weighting, no quantity weighting.  This makes
   the index simple and reproducible, but it does not represent any
   particular household's spending pattern.

2. **Geomean of unequal panels.**  We take an unweighted geomean of
   per-label indices.  Labels with thin panels count the same as labels
   with thousands of rows.  This is appropriate for the intent of the
   exercise (basket-composition sensitivity) but would be wrong for an
   official inflation statistic.

3. **Pattern leakage.**  Karen's regex set is narrow but not perfect.
   Examples surfaced in section 2.4 — e.g., `Tomato (hot house)` falls
   back to a generic `tomato` pattern, dragging in roma tomatoes and
   tomato sauce.  The candidate themes use tighter category patterns but
   are not regression-tested.

4. **Regional skew.**  The parquet is dominated by chains in specific
   regions (e.g., Hannaford in the Northeast).  No basket here corrects
   for that.  All conclusions about the *level* of any index should be
   interpreted as 'representative of the chains we have', not 'national'.

5. **Unit-field is messy.**  The flagged unit-homogeneity issues mostly
   reflect the underlying scrape — many chain pages don't expose a
   structured unit, so we end up with blank `unit` strings.  This is a
   known data-quality issue, not a basket-design problem.  Resolving it
   would involve enriching `unit` with `package_size_raw` parsing (which
   the existing `karens_basket_enriched.csv` already does for a subset).

# Reproducing this report

```bash
# 1. Build the QC table + candidate substitutes + perturbed baskets
python analysis/sandbox/karens_basket_perturbation/build_qc.py

# 2. Compute the basket-level inflation indices
python analysis/sandbox/karens_basket_perturbation/build_indices.py

# 3. Knit this Rmd
Rscript -e "rmarkdown::render('analysis/karens_basket_perturbation.Rmd')"
```

The first two scripts read `data/pricing.parquet` and write CSV artifacts
to `analysis/sandbox/karens_basket_perturbation/`.  This Rmd reads only
those CSV artifacts and never touches the parquet directly, so it knits
quickly.

The three perturbed basket definitions are stored in
`analysis/sandbox/karens_basket_perturbation/perturbed_baskets.csv` with
columns `basket_name, label, patterns_json, source, n_years, n_chains,
n_rows`.  This file is suitable for downstream integration into
`baskets.py` if any of the alternative baskets are adopted as a
permanent, app-visible basket.
