---
title: "County-Level Grocery Price Trend: Statistical Power Analysis"
author: "Grocery Pricing Project"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 6
)
options(scipen = 999)
library(readxl)
library(data.table)
```

# Overview

This report answers a simple question: **when we receive county-level NielsenIQ
Homescan receipt data, which product categories will have enough observations to
support reliable trend analysis of price per unit?**

The approach:

- For each product category, compute the **minimum detectable annual price trend
  (MDE)** at 80% power, as a function of the number of purchase transactions
  observed per county per week.
- Express results as the **minimum viable weekly n** -- the transaction count
  a county-category cell needs to detect a 2%, 5%, or 10% annual price trend.
- Validate the analytical formula against simulation.

## Key assumptions

| Assumption | Value | Notes |
|:-----------|:------|:------|
| Estimand | Annual % change in log(price per unit) | Scale-free; directly interpretable |
| Model | `log(p) ~ week + month_dummies` | OLS with 11 seasonal controls |
| Inference | Sandwich (Newey-West) standard errors | Robust to AR(1) autocorrelation |
| Power target | 80%, two-sided alpha = 0.05 | Standard |
| Autocorrelation | Sweep rho = 0, 0.3, 0.5, 0.7 | Weekly shelf prices are sticky |
| Transaction-level SD | 0.15 / 0.30 / 0.50 | Standardised index / typical / raw mean |
| Time spans | 104, 156, 208, 260 weeks (2-5 years) | |

# Data: NielsenIQ product hierarchy

```{r load-data}
here <- getwd()
# Navigate up if we're in the analysis subfolder
if (basename(here) == "analysis") here <- dirname(here)

hierarchy_path <- file.path(here, "Homescan Disagg Hierarchy.xlsx")

# Read hierarchy
raw <- as.data.table(read_excel(hierarchy_path, sheet = "1-Table-1", skip = 8))
setnames(raw, paste0("V", seq_len(ncol(raw))))
hier <- raw[!is.na(V9), .(
  department    = V5,
  supercategory = V6,
  category      = V7,
  raw_occasions = as.numeric(V9)
)]

# Tag the 20 NHANES-recommended supercategories
# (from nhanes_meal_archetypes analysis -- coverage-maximising selection)
nhanes_picks <- toupper(c(
  "CEREAL & GRANOLA",
  "COFFEE",
  "BREAD",
  "MILK PRODUCTS",
  "EGGS",
  "SALTY SNACKS",
  "COOKIES & CRACKERS",
  "DESSERTS/SWEET GOODS",
  "PREPARED FOODS",
  "MEAL COMBOS",
  "PASTA/RICE/GRAINS",
  "VEGETABLES",
  "FRUIT",
  "CHEESE",
  "BEVERAGES",
  "PIZZA",
  "WRAPS & TORTILLA SHELLS",
  "CONDIMENTS",
  "SEAFOOD",
  "NUT BUTTERS/JAM/JELLIES"
))
hier[, in_selection := toupper(trimws(supercategory)) %in% nhanes_picks]

# Supercategory-level aggregation
sc <- hier[, .(
  n_categories   = .N,
  total_occ      = sum(raw_occasions),
  min_cat_occ    = min(raw_occasions),
  median_cat_occ = median(raw_occasions),
  in_selection   = any(in_selection)
), by = supercategory]
setorder(sc, -total_occ)
```

The Homescan Disagg Hierarchy contains **`r nrow(hier)` product categories**
across **`r uniqueN(hier$supercategory)` supercategories**. Of these,
**`r uniqueN(hier[in_selection == TRUE, supercategory])` supercategories** have
been selected based on the NHANES meal-archetype analysis (a
coverage-maximising 20-category pick designed to reconstruct as many meal
patterns as possible from Homescan purchase data). These cover
**`r nrow(hier[in_selection == TRUE])` categories** and
**`r format(hier[in_selection == TRUE, sum(raw_occasions)], big.mark = ",")`
national purchase occasions** in the latest 52-week period.

```{r hierarchy-table}
sel_sc <- sc[in_selection == TRUE]
knitr::kable(
  sel_sc[, .(
    Supercategory   = supercategory,
    Categories      = n_categories,
    `National Occ.` = format(total_occ, big.mark = ","),
    `Min Cat Occ.`  = format(min_cat_occ, big.mark = ","),
    `Median Cat Occ.` = format(median_cat_occ, big.mark = ",")
  )],
  caption = "NHANES-recommended 20 supercategories, sorted by total national occasions"
)
```

The distribution of national occasions is extremely right-skewed: the median
category has about `r format(round(median(hier$raw_occasions)), big.mark = ",")`
occasions nationally, but the 10th percentile is only
`r round(quantile(hier$raw_occasions, 0.10))`. Once spread across 1,000+
counties, many cells will be thin.

```{r occasions-histogram, fig.height=5}
par(mar = c(5, 4, 3, 1))
cutoff <- 100000L
n_above <- sum(hier$raw_occasions > cutoff)
vals <- hier$raw_occasions[hier$raw_occasions <= cutoff]
h <- hist(vals, breaks = 40, plot = FALSE)
n_bars <- length(h$counts)

bp <- barplot(h$counts, space = 0, col = "#67a9cf", border = "white",
     main = paste0("Distribution of national Raw Occasions per category",
                   "\n(", n_above, " categories above ",
                   formatC(cutoff, format = "d", big.mark = ","), " not shown)"),
     xlab = "Raw Occasions (national, latest 52 weeks)",
     ylab = "Number of categories",
     xaxt = "n")

at_labels <- seq(0, cutoff, by = 20000)
at_pos <- sapply(at_labels, function(v) {
  (v - h$breaks[1]) / diff(range(h$breaks)) * n_bars
})
axis(1, at = at_pos, labels = format(at_labels, big.mark = ","))

# Reference lines: map values to barplot coordinates
med_pos <- (median(hier$raw_occasions) - h$breaks[1]) /
           diff(range(h$breaks)) * n_bars
p10_pos <- (quantile(hier$raw_occasions, 0.10) - h$breaks[1]) /
           diff(range(h$breaks)) * n_bars
abline(v = med_pos, lty = 2, lwd = 2, col = "#b2182b")
abline(v = p10_pos, lty = 3, lwd = 2, col = "#ef8a62")
legend("topright", bty = "n",
       legend = c(
         paste0("Median: ", format(round(median(hier$raw_occasions)), big.mark = ",")),
         paste0("10th pctile: ", round(quantile(hier$raw_occasions, 0.10)))
       ),
       col = c("#b2182b", "#ef8a62"), lty = c(2, 3), lwd = 2)
```

```{r key-findings-compute, include=FALSE}
hier_sel <- hier[in_selection == TRUE]
n_cats_total <- nrow(hier_sel)
n_infeasible_large <- sum(hier_sel$raw_occasions * 0.005 / 52 < 1.6)
n_infeasible_mid   <- sum(hier_sel$raw_occasions * 0.001 / 52 < 1.6)
n_infeasible_small <- sum(hier_sel$raw_occasions * 0.0002 / 52 < 1.6)
```

# Key findings

> **Supercategory-level analysis is viable in most counties.** At the
> supercategory level (e.g. all of VEGETABLES, all of CHEESE), the 20
> NHANES-recommended picks have enough volume for county-level price-trend
> detection in large and mid-size counties with 5 years of data.
>
> **Category-level analysis is not.** When you drill down to individual
> categories (e.g. *Tomatoes*, *Cheddar*, *Potato Chip*), the picture is far
> more sobering:
>
> - **`r n_infeasible_large` of `r n_cats_total` categories (`r round(100*n_infeasible_large/n_cats_total)`%) are infeasible even in large metro counties** -- too few transactions for a single nonzero week.
> - **`r n_infeasible_mid` of `r n_cats_total` (`r round(100*n_infeasible_mid/n_cats_total)`%) are infeasible in mid-size counties.**
> - **`r n_infeasible_small` of `r n_cats_total` (`r round(100*n_infeasible_small/n_cats_total)`%) are infeasible in small/rural counties.**
>
> Only the highest-volume categories (SOFT DRINKS, COWS MILK, BREAD, CHICKEN
> EGGS, POTATOES, POTATO CHIP, and roughly 50--80 others) can support
> category-level county trend analysis. For everything else, you must work at
> the supercategory aggregate or use pooled/hierarchical models.
>
> **The single highest-leverage methodological choice is building a standardised
> price index** (e.g. price-per-ounce or UPC fixed effects) rather than using
> raw category mean prices. This roughly halves the minimum detectable effect
> for every cell.

# Method

## The MDE formula

We model weekly county-category log mean price as:

$$\log(p_w) = \alpha + \beta \cdot w + \sum_{m=2}^{12} \gamma_m \cdot \mathbf{1}[\text{month}=m] + \varepsilon_w$$

where $\varepsilon_w$ follows an AR(1) process with persistence $\rho$.

The OLS slope estimator $\hat\beta$ has true variance under AR(1) errors:

$$\text{Var}(\hat\beta) = \sigma_w^2 \cdot \left[ (X'X)^{-1} X' \Omega X (X'X)^{-1} \right]_{2,2}$$

where $\Omega_{ij} = \rho^{|i-j|}$ is the AR(1) correlation matrix, and
$\sigma_w^2 = \sigma_{\text{txn}}^2 / n + \sigma_{\text{week}}^2$ is the
variance of the weekly log mean price (combining transaction-level variance
averaged down by $n$ transactions, plus irreducible week-level variance).

The **sandwich form** is computed numerically for each $(T, \rho)$ combination
-- this is exact, unlike the commonly used `sqrt((1+rho)/(1-rho))`
approximation which applies only to sample means, not regression slopes with
covariates.

The MDE (annualised) is then:

$$\text{MDE}_{\text{annual}} = 52 \times (t_{\alpha/2, df} + t_{\text{power}, df}) \times \text{SE}_{\text{sandwich}}(\hat\beta)$$

converted from log-points to percentage via $100 \times (e^{\text{MDE}} - 1)$.

## Feasibility filter

Before computing MDEs, we check whether a county-category cell has enough
nonzero weeks. Under a Poisson arrival model, the expected fraction of nonzero
weeks is $1 - e^{-n}$ where $n$ is the expected weekly transaction count. We
require at least 77% of weeks to have $\geq 1$ transaction (annualised: 40 out
of 52 weeks), implying a minimum expected $n \approx 1.6$ per week.

```{r compute-mde}
# --- Parameter grid ---
weekly_n_vals  <- c(1, 2, 5, 10, 25, 50, 100, 250, 500)
T_vals         <- c(104L, 156L, 208L, 260L)
rho_vals       <- c(0.0, 0.3, 0.5, 0.7)
sigma_txn_vals <- c(0.15, 0.30, 0.50)
sigma_week_vals <- c(0.01, 0.03)
target_power   <- 0.80
alpha          <- 0.05

grid <- CJ(
  weekly_n   = weekly_n_vals,
  T_weeks    = T_vals,
  rho        = rho_vals,
  sigma_txn  = sigma_txn_vals,
  sigma_week = sigma_week_vals
)

# Feasibility filter
grid[, nonzero_weeks := T_weeks * (1 - exp(-weekly_n))]
grid[, pct_coverage  := nonzero_weeks / T_weeks]
grid[, feasible := pct_coverage >= 40/52]

# --- Sandwich variance computation ---
sandwich_cache <- list()

get_slope_var_factor <- function(T_weeks, rho) {
  key <- paste(T_weeks, rho, sep = "_")
  if (!is.null(sandwich_cache[[key]])) return(sandwich_cache[[key]])
  
  weeks <- seq_len(T_weeks)
  month_of_week <- ((weeks - 1) %% 52) %/% 4 + 1
  month_of_week[month_of_week > 12] <- 12
  month_mat <- model.matrix(~ factor(month_of_week))[, -1]
  X <- cbind(1, weeks, month_mat)
  
  Omega <- rho^abs(outer(weeks, weeks, "-"))
  XtX_inv <- solve(crossprod(X))
  sandwich <- XtX_inv %*% crossprod(X, Omega %*% X) %*% XtX_inv
  
  result <- list(
    sandwich_var = sandwich[2, 2],
    iid_var      = XtX_inv[2, 2],
    df           = T_weeks - ncol(X)
  )
  sandwich_cache[[key]] <<- result
  return(result)
}

# Pre-populate cache
for (tw in T_vals) for (r in rho_vals) get_slope_var_factor(tw, r)

# --- MDE function ---
compute_mde <- function(weekly_n, T_weeks, rho, sigma_txn, sigma_week,
                         alpha = 0.05, power = 0.80) {
  sigma_w <- sqrt(sigma_txn^2 / weekly_n + sigma_week^2)
  sv <- get_slope_var_factor(T_weeks, rho)
  df <- sv$df
  if (df < 3) return(NA_real_)
  t_alpha <- qt(1 - alpha/2, df)
  t_power <- qt(power, df)
  se_b_true <- sigma_w * sqrt(sv$sandwich_var)
  mde_annual <- 52 * (t_alpha + t_power) * se_b_true
  mde_pct <- 100 * (exp(mde_annual) - 1)
  return(mde_pct)
}

grid[feasible == TRUE, mde_pct := mapply(
  compute_mde, weekly_n, T_weeks, rho, sigma_txn, sigma_week,
  MoreArgs = list(alpha = alpha, power = target_power)
)]

grid[feasible == TRUE, verdict := fcase(
  mde_pct <  2, "well_powered",
  mde_pct <  5, "adequate",
  mde_pct < 10, "marginal",
  mde_pct >= 10, "underpowered",
  default = NA_character_
)]
grid[feasible == FALSE, verdict := "infeasible"]

# --- Min-viable-n lookup ---
thresholds <- c(2, 5, 10)
min_n_list <- list()
for (thresh in thresholds) {
  tmp <- grid[feasible == TRUE & mde_pct < thresh,
              .(min_weekly_n = min(weekly_n)),
              by = .(T_weeks, rho, sigma_txn, sigma_week)]
  tmp[, mde_threshold := thresh]
  min_n_list[[as.character(thresh)]] <- tmp
}
min_n_table <- rbindlist(min_n_list)
setorder(min_n_table, sigma_txn, rho, T_weeks, mde_threshold)
```

# Results

## Five years of data substantially improves power

With 5 years (260 weeks) of receipt data, the model has 247 residual degrees
of freedom after fitting the intercept, trend, and 11 month dummies. This is a
large improvement over 2 years (91 df) -- more weeks means both more precise
slope estimates and better-behaved t-statistics. The tables below show how
minimum-n requirements drop as the time span increases from 2 to 5 years.

## Minimum viable weekly n

The tables below show the minimum average weekly transaction count needed per
county-category cell to detect a given annual price trend, at 80% power. All
use $\sigma_{\text{week}} = 0.01$ (conservative).

### Standardised price index ($\sigma_{\text{txn}} = 0.15$)

A standardised index normalises for package size and brand mix -- the cleanest
measure of actual price movement.

```{r table-std}
make_min_n_display <- function(s_txn, s_week = 0.01) {
  out <- data.table()
  for (tw in T_vals) {
    for (r in rho_vals) {
      vals <- sapply(thresholds, function(th) {
        row <- min_n_table[sigma_txn == s_txn & sigma_week == s_week &
                             T_weeks == tw & rho == r & mde_threshold == th]
        if (nrow(row) == 0) "> 500" else as.character(row$min_weekly_n)
      })
      out <- rbind(out, data.table(
        `Time span` = paste0(tw, " wks (", tw/52, " yr)"),
        `Autocorrelation (rho)` = r,
        `MDE < 2%/yr` = vals[1],
        `MDE < 5%/yr` = vals[2],
        `MDE < 10%/yr` = vals[3]
      ))
    }
  }
  return(out)
}
knitr::kable(make_min_n_display(0.15),
             caption = "Minimum weekly n: standardised price index (sigma_txn = 0.15)")
```

### Typical category ($\sigma_{\text{txn}} = 0.30$)

Representative of a reasonably well-defined product category or a decent price
index with some residual mix noise.

```{r table-typ}
knitr::kable(make_min_n_display(0.30),
             caption = "Minimum weekly n: typical category (sigma_txn = 0.30)")
```

### Raw category mean ($\sigma_{\text{txn}} = 0.50$)

If you use the unweighted average receipt price across all UPCs in a category,
package-size and brand mixing inflate variance substantially.

```{r table-raw}
knitr::kable(make_min_n_display(0.50),
             caption = "Minimum weekly n: raw category mean (sigma_txn = 0.50)")
```

**Reading these tables:** at the intersection of your time span, autocorrelation
assumption, and desired MDE threshold, the number shown is the minimum average
weekly transaction count a county-category cell needs. For example, with 5 years
of data, moderate autocorrelation ($\rho = 0.5$), and a typical category
($\sigma = 0.30$), look up the row for 260 wks / rho = 0.5 to find the
required weekly n for each MDE threshold.

## Category feasibility: NHANES-recommended 20

The tables below map these thresholds to the 20 NHANES-recommended
supercategories using three illustrative county-size scenarios: **large metro**
(0.5% of the national panel), **mid-size** (0.1%), and **small/rural** (0.02%).
All assume **5 years of data** (260 weeks) and moderate autocorrelation
($\rho = 0.5$).

```{r feasibility-tables, results='asis'}
shares       <- c(0.005, 0.001, 0.0002)
share_labels <- c("Large metro (0.5%)", "Mid-size (0.1%)", "Small/rural (0.02%)")

for (s_txn in c(0.15, 0.30, 0.50)) {
  label <- if (s_txn == 0.15) "Standardised index (sigma = 0.15)" else
           if (s_txn == 0.30) "Typical category (sigma = 0.30)" else
                              "Raw category mean (sigma = 0.50)"
  
  rows <- list()
  for (i in seq_len(nrow(sel_sc))) {
    row <- sel_sc[i]
    vals <- character(3)
    for (j in seq_along(shares)) {
      wn <- row$total_occ * shares[j] / 52
      if (wn < 1.6) {
        vals[j] <- paste0("n=", round(wn, 1), " INFEASIBLE")
      } else {
        m <- compute_mde(weekly_n = wn, T_weeks = 260, rho = 0.5,
                         sigma_txn = s_txn, sigma_week = 0.01)
        verd <- if (is.na(m)) "NA" else
                if (m < 2)  "well_powered" else
                if (m < 5)  "adequate" else
                if (m < 10) "marginal" else "underpowered"
        vals[j] <- paste0(sprintf("%.1f%%", m), " ", verd)
      }
    }
    rows[[i]] <- data.table(
      Supercategory = row$supercategory,
      `Natl Occasions` = format(row$total_occ, big.mark = ","),
      V1 = vals[1], V2 = vals[2], V3 = vals[3]
    )
  }
  tbl <- rbindlist(rows)
  setnames(tbl, c("V1","V2","V3"), share_labels)
  
  cat("\n\n###", label, "\n\n")
  print(knitr::kable(tbl,
    caption = paste0("Feasibility at rho=0.5, T=260 weeks (5 yr): ", label)))
  cat("\n")
}
```

## Category-level feasibility

The supercategory-level picture is encouraging, but most research questions
target specific product categories (e.g. *Tomatoes*, *Cheddar*, *Potato Chip*),
not broad aggregates like VEGETABLES or CHEESE. There are **`r nrow(hier[in_selection == TRUE])` categories** within the 20 NHANES-recommended supercategories, and many of them are far too thin for county-level work.

```{r cat-level-compute}
# Category-level feasibility at central scenario: T=260, rho=0.5, sigma_week=0.01
cat_data <- hier[in_selection == TRUE, .(supercategory, category, raw_occasions)]
setorder(cat_data, -raw_occasions)

shares       <- c(0.005, 0.001, 0.0002)
share_labels <- c("Large metro (0.5%)", "Mid-size (0.1%)", "Small/rural (0.02%)")
sigma_txn_scenarios <- c(0.15, 0.30, 0.50)
sigma_labels <- c("0.15" = "Std index", "0.3" = "Typical", "0.5" = "Raw mean")

# Compute MDE for every category x county-tier x sigma_txn combo
cat_results <- list()
for (s_txn in sigma_txn_scenarios) {
  for (j in seq_along(shares)) {
    tmp <- copy(cat_data)
    tmp[, weekly_n := raw_occasions * shares[j] / 52]
    tmp[, feasible := weekly_n >= 1.6]
    tmp[feasible == TRUE, mde_pct := sapply(weekly_n, function(wn) {
      compute_mde(wn, T_weeks = 260, rho = 0.5,
                  sigma_txn = s_txn, sigma_week = 0.01)
    })]
    tmp[feasible == TRUE, verdict := fcase(
      mde_pct <  2, "well_powered",
      mde_pct <  5, "adequate",
      mde_pct < 10, "marginal",
      mde_pct >= 10, "underpowered"
    )]
    tmp[feasible == FALSE, verdict := "infeasible"]
    tmp[, sigma_txn := s_txn]
    tmp[, county_tier := share_labels[j]]
    cat_results[[length(cat_results) + 1]] <- tmp
  }
}
cat_all <- rbindlist(cat_results)
```

### Summary: how many categories survive?

```{r cat-summary-table, results='asis'}
verdict_order <- c("well_powered", "adequate", "marginal", "underpowered", "infeasible")

for (s_txn in sigma_txn_scenarios) {
  label <- sigma_labels[as.character(s_txn)]
  sub <- cat_all[sigma_txn == s_txn]
  
  summ <- sub[, .N, by = .(county_tier, verdict)]
  summ[, verdict := factor(verdict, levels = verdict_order)]
  wide <- dcast(summ, county_tier ~ verdict, value.var = "N", fill = 0,
                drop = FALSE)
  # Ensure all verdict columns exist
  for (v in verdict_order) {
    if (!v %in% names(wide)) wide[, (v) := 0L]
  }
  # Reorder rows
  wide[, county_tier := factor(county_tier, levels = share_labels)]
  setorder(wide, county_tier)
  wide[, Total := rowSums(.SD), .SDcols = verdict_order]
  wide[, `Viable (MDE<10%)` := rowSums(.SD),
       .SDcols = c("well_powered", "adequate", "marginal")]
  
  cat("\n\n####", paste0(label, " (sigma_txn = ", s_txn, ")"), "\n\n")
  print(knitr::kable(wide,
    caption = paste0("Category-level verdict counts (T=260 wks, rho=0.5): ", label)))
  cat("\n")
}
```

At **category level**, the picture is much more sobering than at supercategory
level. Over half of the 381 categories are infeasible in mid-size counties, and
the vast majority are infeasible in small/rural counties regardless of the
variance scenario.

### Distribution of category-level MDEs

```{r cat-mde-histogram, fig.height=7}
par(mfrow = c(3, 1), mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))

for (s_txn in sigma_txn_scenarios) {
  label <- sigma_labels[as.character(s_txn)]
  # Show large metro (best case) — feasible categories only
  sub <- cat_all[sigma_txn == s_txn & county_tier == share_labels[1] &
                   feasible == TRUE & !is.na(mde_pct)]
  n_infeasible <- cat_all[sigma_txn == s_txn & county_tier == share_labels[1] &
                            feasible == FALSE, .N]
  
  if (nrow(sub) > 0) {
    vals <- pmin(sub$mde_pct, 30)  # truncate for display
    h <- hist(vals, breaks = 30, plot = FALSE)
    n_bars <- length(h$counts)
    barplot(h$counts, space = 0, col = "#67a9cf", border = "white",
            main = paste0(label, " -- Large metro counties",
                          "\n(", n_infeasible, " of ",
                          nrow(cat_data), " categories infeasible, not shown)"),
            ylab = "Number of categories",
            xlab = "MDE (% per year)",
            xaxt = "n")
    at_labels <- seq(0, 30, by = 5)
    at_pos <- sapply(at_labels, function(v) {
      (v - h$breaks[1]) / diff(range(h$breaks)) * n_bars
    })
    axis(1, at = at_pos, labels = paste0(at_labels, "%"))
    # Threshold lines
    for (thresh in c(2, 5, 10)) {
      thresh_pos <- (thresh - h$breaks[1]) / diff(range(h$breaks)) * n_bars
      abline(v = thresh_pos, lty = 2, lwd = 2, col = "grey40")
    }
  }
}
mtext("Category-level MDE distribution (large metro, T=260 wks, rho=0.5)",
      outer = TRUE, cex = 1.1, font = 2)
```

### Which supercategories can support category-level analysis?

The table below shows, for each supercategory, how many of its constituent
categories are viable (MDE < 10%/yr) at each county tier.

```{r cat-thinning-table, results='asis'}
for (s_txn in sigma_txn_scenarios) {
  label <- sigma_labels[as.character(s_txn)]
  sub <- cat_all[sigma_txn == s_txn]
  
  thin <- sub[, .(
    total    = .N,
    viable   = sum(verdict %in% c("well_powered", "adequate", "marginal"))
  ), by = .(supercategory, county_tier)]
  
  thin[, pct_viable := sprintf("%d/%d", viable, total)]
  wide <- dcast(thin, supercategory ~ county_tier, value.var = "pct_viable")
  
  # Add total occasions for sorting
  wide <- merge(wide, sc[, .(supercategory, total_occ)], by = "supercategory")
  setorder(wide, -total_occ)
  wide[, total_occ := NULL]
  
  # Reorder columns
  setcolorder(wide, c("supercategory", share_labels))
  
  cat("\n\n####", paste0(label, " (sigma_txn = ", s_txn, ")"), "\n\n")
  print(knitr::kable(wide,
    caption = paste0("Viable categories per supercategory (MDE<10%, T=260 wks, rho=0.5): ",
                     label),
    col.names = c("Supercategory", share_labels)))
  cat("\n")
}
```

### Top viable categories in mid-size counties

The following table lists categories that achieve MDE < 10%/yr in a mid-size
county (0.1% of national panel), across all three variance scenarios. These are
the categories where county-level trend analysis is most realistic.

```{r cat-detail-table, results='asis'}
for (s_txn in sigma_txn_scenarios) {
  label <- sigma_labels[as.character(s_txn)]
  
  rows <- list()
  for (i in seq_len(nrow(cat_data))) {
    row <- cat_data[i]
    vals <- character(3)
    for (j in seq_along(shares)) {
      wn <- row$raw_occasions * shares[j] / 52
      if (wn < 1.6) {
        vals[j] <- "INFEASIBLE"
      } else {
        m <- compute_mde(weekly_n = wn, T_weeks = 260, rho = 0.5,
                         sigma_txn = s_txn, sigma_week = 0.01)
        verd <- if (is.na(m)) "NA" else
                if (m < 2)  "well_powered" else
                if (m < 5)  "adequate" else
                if (m < 10) "marginal" else "underpowered"
        vals[j] <- paste0(sprintf("%.1f%%", m), " ", verd)
      }
    }
    rows[[i]] <- data.table(
      Supercategory = row$supercategory,
      Category = row$category,
      `Natl Occ` = format(row$raw_occasions, big.mark = ","),
      V1 = vals[1], V2 = vals[2], V3 = vals[3]
    )
  }
  tbl <- rbindlist(rows)
  setnames(tbl, c("V1","V2","V3"), share_labels)
  
  # Filter to categories viable in mid-size county
  tbl_viable <- tbl[!grepl("INFEASIBLE|underpowered", `Mid-size (0.1%)`)]
  
  cat("\n\n####", paste0(label, " (sigma_txn = ", s_txn, ")"), "\n\n")
  if (nrow(tbl_viable) == 0) {
    cat("*No categories achieve MDE < 10%/yr in mid-size counties at this variance level.*\n")
  } else {
    print(knitr::kable(tbl_viable,
      caption = paste0("Categories viable in mid-size counties (MDE<10%, T=260 wks, rho=0.5): ",
                       label)))
  }
  cat("\n")
}
```

## MDE curves

The plots below show how the minimum detectable effect (annual %) decreases as
weekly transaction count increases, for each time span and autocorrelation
level. The dashed lines mark the 2%, 5%, and 10% MDE thresholds. Solid lines
show 2 years, dashed lines 5 years -- illustrating the power gain from the full
data window.

```{r mde-curves, fig.height=8}
# Compute MDE on a finer n grid for plotting
plot_n <- seq(2, 500, by = 1)
plot_grid <- CJ(
  weekly_n  = plot_n,
  T_weeks   = T_vals,
  rho       = rho_vals,
  sigma_txn = c(0.15, 0.30, 0.50)
)
plot_grid[, mde_pct := mapply(
  compute_mde, weekly_n, T_weeks, rho, sigma_txn,
  MoreArgs = list(sigma_week = 0.01, alpha = alpha, power = target_power)
)]

# Limit display range
plot_grid <- plot_grid[!is.na(mde_pct) & mde_pct < 50]

sigma_labels <- c("0.15" = "Standardised index",
                   "0.3"  = "Typical category",
                   "0.5"  = "Raw category mean")

par(mfrow = c(3, 1), mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
rho_cols <- c("0"   = "#2166ac",
              "0.3" = "#67a9cf",
              "0.5" = "#ef8a62",
              "0.7" = "#b2182b")

for (s_txn in c(0.15, 0.30, 0.50)) {
  sub <- plot_grid[sigma_txn == s_txn & T_weeks %in% c(104, 260)]
  
  plot(NULL, xlim = c(2, 300), ylim = c(0, 25),
       xlab = "Weekly transactions per county-category",
       ylab = "MDE (% per year)",
       main = sigma_labels[as.character(s_txn)])
  abline(h = c(2, 5, 10), lty = 2, col = "grey60")
  text(290, c(2, 5, 10) + 0.6, c("2%", "5%", "10%"),
       cex = 0.8, col = "grey40")
  
  for (tw in c(104, 260)) {
    for (r in rho_vals) {
      d <- sub[T_weeks == tw & rho == r]
      if (nrow(d) == 0) next
      lty_val <- if (tw == 104) 1 else 2
      lines(d$weekly_n, d$mde_pct,
            col = rho_cols[as.character(r)], lwd = 1.5, lty = lty_val)
    }
  }
  
  legend("topright", bty = "n", cex = 0.75, ncol = 2,
         legend = c(paste0("rho=", names(rho_cols)), "2 years", "5 years"),
         col = c(rho_cols, "black", "black"),
         lty = c(rep(1, 4), 1, 2),
         lwd = 1.5)
}
mtext("MDE vs. weekly transaction count (sigma_week = 0.01)", outer = TRUE,
      cex = 1.1, font = 2)
```

# Simulation validation

To confirm the analytical formula is correctly calibrated, we simulate 1,000
datasets at the analytical MDE and check that the empirical rejection rate
matches 80%.

```{r simulation, cache=TRUE}
set.seed(42)
n_sim <- 1000L

sim_combos <- data.table(
  weekly_n   = c(10,   50,  100,  25),
  T_weeks    = c(104, 208,  260, 156),
  rho        = c(0.0, 0.5,  0.7, 0.3),
  sigma_txn  = c(0.30, 0.30, 0.15, 0.50),
  sigma_week = c(0.01, 0.01, 0.01, 0.03)
)

sim_results <- list()

for (sc_i in seq_len(nrow(sim_combos))) {
  p <- sim_combos[sc_i]
  
  sigma_w <- sqrt(p$sigma_txn^2 / p$weekly_n + p$sigma_week^2)
  sv <- get_slope_var_factor(p$T_weeks, p$rho)
  df <- sv$df
  t_a <- qt(1 - alpha/2, df)
  t_p <- qt(target_power, df)
  se_b_true <- sigma_w * sqrt(sv$sandwich_var)
  mde_slope <- (t_a + t_p) * se_b_true
  
  weeks <- seq_len(p$T_weeks)
  month_of_week <- ((weeks - 1) %% 52) %/% 4 + 1
  month_of_week[month_of_week > 12] <- 12
  month_mat <- model.matrix(~ factor(month_of_week))[, -1]
  Xfull <- cbind(1, weeks, month_mat)
  XtX_inv <- solve(crossprod(Xfull))
  Omega_sim <- (p$rho)^abs(outer(weeks, weeks, "-"))
  
  rejections <- 0L
  for (sim in seq_len(n_sim)) {
    eps <- numeric(p$T_weeks)
    eps[1] <- rnorm(1, 0, sigma_w)
    for (t in 2:p$T_weeks) {
      eps[t] <- p$rho * eps[t-1] + rnorm(1, 0, sigma_w * sqrt(1 - p$rho^2))
    }
    seasonal <- 0.02 * sin(2 * pi * weeks / 52)
    y <- mde_slope * weeks + seasonal + eps
    
    fit <- lm.fit(Xfull, y)
    coefs <- fit$coefficients
    resids <- fit$residuals
    sigma_hat <- sqrt(sum(resids^2) / df)
    sandwich_se <- sigma_hat * sqrt(
      (XtX_inv %*% crossprod(Xfull, Omega_sim %*% Xfull) %*% XtX_inv)[2, 2]
    )
    t_stat <- coefs[2] / sandwich_se
    if (abs(t_stat) > qt(1 - alpha/2, df)) rejections <- rejections + 1L
  }
  
  sim_results[[sc_i]] <- data.table(
    `Weekly n`    = p$weekly_n,
    `T (weeks)`   = p$T_weeks,
    `rho`         = p$rho,
    `sigma_txn`   = p$sigma_txn,
    `sigma_week`  = p$sigma_week,
    `Analytical MDE (%/yr)` = sprintf("%.1f%%",
                                       100 * (exp(52 * mde_slope) - 1)),
    `Empirical power` = sprintf("%.1f%%", 100 * rejections / n_sim),
    `Target`      = "80%"
  )
}

knitr::kable(rbindlist(sim_results),
             caption = "Simulation validation: empirical power at the analytical MDE (1,000 replications)")
```

Empirical power is consistently near 80% (small upward bias is expected from
the conservative sandwich formula), confirming the analytical MDE is correctly
calibrated across the 2--5 year range.

# Recommendations

1. **Five years of data is a major advantage.** Compared to 2 years, the 5-year
   window roughly halves the minimum-n requirements for trend detection. Many
   categories that would be marginal at 2 years become adequate or well-powered
   at 5 years.

2. **Invest in a standardised price index.** The difference between raw
   category means ($\sigma = 0.50$) and a fixed-weight index ($\sigma = 0.15$)
   shifts a county-category cell from "marginal" to "well-powered." This is the
   single highest-leverage methodological choice after time span.

3. **Supercategory-level analysis is broadly viable; category-level is not.**
   At supercategory level, most of the 20 NHANES-recommended picks are adequate
   or better in large and mid-size counties. But at category level, the majority
   of the 381 constituent categories are infeasible or underpowered in all but
   the largest metro counties. County-level trend analysis at category
   granularity is realistic only for the highest-volume categories (e.g. SOFT
   DRINKS, COWS MILK, BREAD, CHICKEN EGGS, POTATOES, POTATO CHIP).

4. **For category-level questions in smaller counties, aggregate or pool.**
   Options include: (a) work at the supercategory level, (b) use a hierarchical
   model that borrows strength across counties, or (c) group related categories
   (e.g. all chip types into a single "chips" aggregate). This power analysis
   assumes standalone county-by-county regressions -- the most conservative
   framing.

5. **When the actual county data arrives**, compute each county's average weekly
   $n$ per category and look it up in the minimum-viable-n tables to get
   immediate go/no-go verdicts. The category-level thinning tables in this
   report identify which supercategories can support category-level work and
   which cannot.

# Technical notes

- The sandwich variance $(X'X)^{-1} X' \Omega X (X'X)^{-1}$ is computed
  exactly for each $(T, \rho)$ combination rather than using the
  $\sqrt{(1+\rho)/(1-\rho)}$ approximation, which applies only to sample means
  and substantially misestimates the OLS slope variance in the presence of
  covariates.
- The feasibility filter uses a Poisson model for zero-week probability: a cell
  with expected weekly $n < 1.6$ will have $>23\%$ of weeks with no
  transactions and is marked infeasible.
- Verdict bands: **well-powered** (MDE < 2%/yr), **adequate** (2--5%),
  **marginal** (5--10%), **underpowered** (> 10%). These are calibrated to US
  grocery inflation norms (typically 2--4%/yr in non-crisis periods).
- All code and output CSVs are in `analysis/`.

```{r write-csv}
out_dir <- file.path(here, "analysis")
fwrite(grid, file.path(out_dir, "power_mde_county_trends.csv"))
fwrite(min_n_table, file.path(out_dir, "power_min_viable_n.csv"))
```

---

*Output files: `power_mde_county_trends.csv` (`r nrow(grid)` rows),
`power_min_viable_n.csv` (`r nrow(min_n_table)` rows)*
